Urban and socio-economic correlates of property price in Dublin's area

### 2020-08-18 manuscript/manuscript.tex
reviewer changes

\hypersetup{
            pdftitle={Urban and socio-economic correlates of property price evolution:~Application to Dublin's costal area},
            pdfkeywords={Property price, Housing Market, Feature Analysis, Machine Learning, Geocoding},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Urban and socio-economic correlates of property price evolution:~Application to Dublin's costal area}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Understanding the characteristics of the housing market is essential for both sellers and buyers. However, the housing market is influenced by multiple factors. In this paper, the urban and socio-economic structure of an area is used to predict the price of 10389 houses sold in 2018 in the city of Dublin. More precisely, the direct distance from each house to 160 urban features taken from OpenStreetMap is calculated, and an extreme gradient boosting linear regression performed. Using these features, the model explains 47\% of the housing price variance. The most important features in this model are the proximity to an embassy and to a grassland. In addition, the results of a population census from 2016 are also used to correlate with the price of houses. From this census, 48 features are used as the input of a gradient boost linear regression model. In all, the socio-economic features are explaining 42\% of the housing price variance as well. The most important socio-economic features are; the density of houses having more than eight or more rooms in the area, the density of young children, and the density of individuals reporting that they have no religion. By taking into account either urban or socio-economic features, it is possible to accurately estimate housing prices and to predict their evolution.
\end{abstract}

@@

Beside their inherent characteristics such as size, design and materials (Bourassa, Cantoni, and Hoesli 2007; Liu 2013), property valuation is dictated by multiple external factors explaining their spatial autocorrelation (Basu and Thibodeau 1998). Some of these factors are macroeconomic (Antonakakis and Floros 2016), related to the evolution of population general wealth while others are related to the characteristics of the neighbourhood (Dubin 1998). These characteristics include the urban and socio economic structure of the neighbourhood (Goodman and Thibodeau 1998). However, it is difficult to evaluate the influence of such external factors on property valuation (Clapp, Kim, and Gelfand 2002). By using machine learning algorithms, prediction of house prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market (Limsombunchai 2004; Feng and Jones 2015; Selim 2009).

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customisation and high speed.

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes (Authority 2020). Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed.

Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only houses sold in 2018 were processed to avoid potential temporal incoherences. In 2018, a total of 10395 property in the area of Dublin were sold, and these constitute the database used to identify the urban and socio-economic correlates of property prices.

In order to evaluate the spacial distribution of the property sold, property addresses were geocoded (i.e., converted to latitude and longitude) using the OpenStreetMap API. OpenStreetMap is a collaborative project which aims to create and provide access to free editable maps of the world (Haklay and Weber 2008).

To estimate the geographical distribution of the price density, multiple methods such as multiple regression (McCluskey et al. 2000) or Bayesian smoothing (Clapp, Kim, and Gelfand 2002) have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. To deal with the influence of boundaries on price estimation, a generalised additive model using soap film smoothing has been used (Wood, Bravington, and Hedley 2008).

@@

OpenStreetMap combines information about more than 400 urban features including road information and building information to categorize features such as amenities, leisure or tourism structures (OpenStreetMap 2020). Among these features, only 160 are available or relevant to the Dublin Area. The distance between each property and the closest point corresponds to each of the 160 urban features.

@@

In addition to the distance to every urban feature, each house sold in 2018 was related to 48 socio-economic features of the small area including the house, as measured by the Irish census 2016 (Office 2020). The results of Irish 2016 census consultation is accessible through both the Central Statistics Office and All-Island Research Observatory. The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

\hypertarget{gradient-boosting-regressions}{%
\subsection{Gradient Boosting Regressions}\label{gradient-boosting-regressions}}

Lastly, an eXtreme Gradient Boosting (XGBoost) regression model was calculated to examine the extent to which house prices rely on urban and on socio-economic features. The XGBoost algorithm optimizes the prediction accuracy by performing iterative least-squares regressions (set to 100 iterations as the best trade-off between accuracy and speed), thereby minimizing the root mean squared error (Friedman 2001; Chen and Guestrin 2016). The original dataset has been randomly split to 80\% for training and 20\% for testing the models. Models accuracy is estimated using the coefficient of determination (\(R^2\)), the Root Mean Square Error (\(RMSE\)) and the Mean Absolute Error (\(MAE\)) of the predicted property price values of the test dataset.

@@

In 2018, the average price of a sold property in Dublin was €330,364 (SD = €180,448). In order to remove potential human errors and outliers, prices higher or lower than 1 SD were removed from the original dataset.

\begin{figure*}
\includegraphics[width=0.98\textwidth]{dsaa_application_manuscript_files/figure-latex/distrib-plot-1} \caption{Overall (A) and geographic (B) distribution of the houses prices density in the city of Dublin in 2018. The geographic distribution was obtained by using a generalized additive model with soap film smooth parameter to take into account the influence of coastal boundaries. Geographic estimates outside of the 95\% Confidence Interval were removed.}\label{fig:distrib-plot}
\end{figure*}

The density of housing prices reveals a unimodal distribution of property prices (Figure \ref{fig:distrib-plot}A). In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \ref{fig:distrib-plot}B).

@@

Using the XGBoost algorithm, the 160 urban features are explaining 44\% of the property price variance (\(F(1, 2076) = 1,629.56\), \(p < .001\)) with a \(RMSE\) of €127,783 and a \(MAE\) of €85,115.25.

\begin{figure}
\includegraphics[width=0.98\columnwidth]{dsaa_application_manuscript_files/figure-latex/osm-features-xgb-1} \caption{Property price prediction accuracy (A) and Property price prediction error (B) using urban features with XGBoost.}\label{fig:osm-features-xgb}
\end{figure}

The overall correlation of the predicted property prices with urban features is shown in Figure \ref{fig:osm-features-xgb}A. It can be noticed that property prices situated at the low and the high end of the range are the most difficult to predict (Figure \ref{fig:osm-features-xgb}B). A possible explanation is the absence of distinctive and recurrent patterns in urban features for these houses. However, the prices higher than €300,000 are potentially driving down the prediction accuracy due to outliers.

\begin{table}

\caption{\label{tab:osm-features-table}Urban features importance (higher than 1\%).}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
amenity & embassy & 17.7\%\\
natural & grassland & 6.7\%\\
route & bus & 2.3\%\\
power & line & 2.2\%\\
boundary & administrative & 1.5\%\\
place & island & 1.5\%\\
barrier & wall & 1.5\%\\
cycleway & share & 1.4\%\\
route & ferry & 1.3\%\\
cutting & yes & 1.1\%\\
amenity & bar & 1.1\%\\
barrier & yes & 1.1\%\\
place & locality & 1.1\%\\
boundary & political & 1.1\%\\
route & road & 1.1\%\\
highway & path & 1.1\%\\
tunnel & yes & 1.0\%\\
boundary & postal & 1.0\%\\
\bottomrule
\end{tabu}
\end{table}

By analysing their importance (Table \ref{tab:osm-features-table}), the most relevant geographical features to predict housing prices are the distance to an embassy (17.7\%) and the distance to natural grasslands such as parks and gardens (6.7\%).

@@

Using the XGBoost algorithm, the 48 socio-economic features explain 42\% of the property price variance (\(F(1, 2076) = 1,500.87\), \(p < .001\)) with a \(RMSE\) of €131,270 and a \(MAE\) of €86,891.85.

\begin{figure}
\includegraphics[width=0.98\columnwidth]{dsaa_application_manuscript_files/figure-latex/census-features-xgb-1} \caption{Property price prediction accuracy (A) and Property price prediction error (B) using socio-economic features with XGBoost.}\label{fig:census-features-xgb}
\end{figure}

@@

\begin{table}

\caption{\label{tab:census-features-table}Socio-economic features importance (higher than 1\%).}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lll}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
housing rooms & 8 or more Rooms & 28.2\%\\
population & Age 0 - 14 & 4.5\%\\
religion & No Religion & 4.4\%\\
population & Age 80 Plus & 3.6\%\\
general health & Very Good & 3.5\%\\
housing rooms & 6 Rooms & 3.2\%\\
housing tenure & Owner Occupier with Mortgage & 2.5\%\\
disabilty age group & Persons with a disability aged 65 Plus & 2.3\%\\
housing rooms & 5 Rooms & 2.2\%\\
housing tenure & Social Rented & 2.2\%\\
housing rooms & 4 Rooms & 2.1\%\\
religion & Other Catholic & 2.0\%\\
housing rooms & 7 Rooms & 1.9\%\\
housing rooms & 3 Rooms & 1.8\%\\
religion & No Answer & 1.8\%\\
housing tenure & Private Rented & 1.6\%\\
population & Age 45 - 64 & 1.5\%\\
population & Age 25 - 44 & 1.5\%\\
housing tenure & Owner Occupier No Mortgage & 1.5\%\\
disabilty age group & Persons with a disability aged 0 - 14 & 1.5\%\\
disabilty age group & Persons with a disability aged 25 - 44 & 1.4\%\\
housing rooms & 1 Room & 1.4\%\\
carers & Provides No Care & 1.3\%\\
religion & Roman Catholic & 1.3\%\\
general health & Good & 1.3\%\\
population & Female & 1.3\%\\
general health & Fair & 1.3\%\\
disabilty age group & Total persons with a disability & 1.2\%\\
general health & Bad & 1.2\%\\
religion & Other Religion & 1.2\%\\
disabilty age group & Persons with a disability aged 45 - 64 & 1.2\%\\
carers & Total Care Providers & 1.2\%\\
population & Age 15 - 24 & 1.1\%\\
carers & 20-49 hours unpaid PW & 1.1\%\\
housing tenure & Rented Free of Rent & 1.0\%\\
\bottomrule
\end{tabular}
\end{table}

According to the analysis of the relative importance of socio-economic features (Table \ref{tab:census-features-table}), the most important socio-economic features are the proportion of large houses (i.e., houses with eight or more rooms) in the small area containing the property (28.2\%). It also appears that areas having a high proportion of young children (4.4\%), as well as the proportion of people reporting that they have no religion (4.5\%), influence the model.

@@

While Dublin is a very specific city due to its low population density spread on a large surface, its housing market is one of the most expensive in Europe. The main reason is the structure of the market, mainly made of residential properties and few apartment buildings. In this study, the prices of 10387 properties sold in 2018 in the area of Dublin were analysed from the Property Services Regulatory Authority database in order to identify the potential factors correlating with these prices. The enforcement of a housing price registry has been shown as being an essential tool to regulate the housing market (Tomson 2016). However, a major limitation of this study is the absence of characteristics for the property itself. Indeed, prices are mainly determined by factors such as the size of the property or the number of rooms.

Despite the absence of property characteristics, urban and socio-economic features were used to identify spatial correlates to the housing prices. Results revealed that proximity to embassies and grasslands is a driver of house prices. Embassies are in general located in the most expensive areas of cities which are also the most aesthetically pleasing and the most secure part of cities. In addition, the density of embassies and properties allocated to embassy staff reduces the density of property available on the market and drives their prices up. The distance to a park or a natural area is also a very important factor for house prices. Results also revealed that the density of large houses in the area and the proportion of individuals reporting having no religion are very important. Again, the size of the property sold was not included in the database; however, the higher the density of large houses in the area, the higher the probability for the property to be a large house. The second most important feature is the density of young children. While the literature is not conclusive of the link between wealth and family size, future research could investigate the mediation effect of house size between family size and property price. Finally, the relation with the expression of religious belief, or more precisely its absence, is a very interesting feature. Again, the relationship between religion and income is not confirmed by the literature. Whereas some studies reveal a positive correlation between religion and income (Guiso, Sapienza, and Zingales 2003; Elgin et al. 2013) others reveals no evidence of such relationship (De La O and Rodden 2008). Here, it appears that prices of houses are higher in areas with a high density of people indicating that they have no religion.

@@

The evolution of housing prices is a pressing issue in most European capitals and specially in Dublin. Given their significant increase, houses are less and less affordable for individuals. By preforming a feature analysis with urban and socio-economic features, it is possible to evaluate and predict the potential price of a house. Indeed features such as the presence of embassies or parks are criteria that influence significantly the price of houses. Similarly, the characteristics of inhabitants in the area such as religion, health and age is correlated to the evolution of housing prices. These results allow an understanding of why some areas have higher prices than others.

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-antonakakis2016dynamic}{}%
Antonakakis, Nikolaos, and Christos Floros. 2016. ``Dynamic Interdependencies Among the Housing Market, Stock Market, Policy Uncertainty and the Macroeconomy in the United Kingdom.'' \emph{International Review of Financial Analysis} 44: 111--22.

@@

\leavevmode\hypertarget{ref-rppr2020}{}%
Authority, Property Services Regulatory. 2020. ``Residential Property Price Register.'' June 1, 2020. \url{https://propertypriceregister.ie}.

@@

@@

@@

>>>>

\hypersetup{
            pdftitle={Urban and socio-economic correlates of property price in Dublin's area},
            pdfkeywords={Property price, Housing Market, Feature Analysis, Machine Learning, Geocoding},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{graphicx}
\usepackage{flushend}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Urban and socio-economic correlates of property price in Dublin's area}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Understanding the characteristics of the housing market is essential for both sellers and buyers. However, the housing market is influenced by multiple factors. In this paper, the urban and socio-economic structure of an area is used to predict the price of 10387 properties sold in 2018 in the city of Dublin. More precisely, the direct distance from each property to 160 urban features taken from OpenStreetMap is calculated, and an extreme gradient boosting linear regression performed. Using these features, the model explains 45\% of the housing price variance. The most important features in this model are the proximity to an embassy and to a grassland. In addition, the results of a population census from 2016 are also used to correlate with the price of properties. From this census, 48 features are used as the input of a gradient boost linear regression model. In all, the socio-economic features are explaining 43\% of the housing price variance as well. The density of individuals reporting that they are not providing unpaid personal help for a friend or family member as well as individuals reporting that they have no religion are the most important socio-economic feratures. By taking into account either urban or socio-economic features, it is possible to accurately estimate housing prices and to predict their evolution.
\end{abstract}

@@

Beside their inherent characteristics such as size, design and materials (Bourassa, Cantoni, and Hoesli 2007; Liu 2013), property valuation is dictated by multiple external factors explaining their spatial autocorrelation (Basu and Thibodeau 1998). Some of these factors are macroeconomic (Antonakakis and Floros 2016), related to the evolution of population general wealth while others are related to the characteristics of the neighbourhood (Dubin 1998). These characteristics include the urban and socio economic structure of the neighbourhood (Goodman and Thibodeau 1998). However, it is difficult to evaluate the influence of such external factors on property valuation (Clapp, Kim, and Gelfand 2002). By using machine learning algorithms, prediction of property prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market (Limsombunchai 2004; Feng and Jones 2015; Selim 2009). Indeed, artificial neural network techniques can quickly adjust to volatility changes, therefore incorporating sharp increases or decreases in their predictions (Aragonés, Blanco, and Estévez 2007). However, the model obtained from most artificial neural networks are hardly interpretable (Rudin 2019).

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customisation and high speed (Chen and Guestrin 2016). Contrary to most artificial neural network techniques, XGBoost is an interpretation-focused method which automatically provides estimates of feature importance from a trained predictive model.

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes (Property Services Regulatory Authority 2020). Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed.

Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. The total amount of property sold in 2018 in the county Dublin which appears in the Property Price Register is 18665. From these, all property corresponding to apartments were removed (10.3\% of the original dataset). Finally, unsuccessfully geocoded properties were also removed from the original dataset (38.0\%) to obtain the 10389 of the final dataset used to identify the urban and socio-economic correlates of property prices.

In order to evaluate the spacial distribution of the property sold, property addresses were geocoded (i.e., converted to latitude and longitude) using the Nominatim API (Hoffmann 2020) to OpenStreetMap data. OpenStreetMap is a collaborative project which aims to create and provide access to free editable maps of the world (Haklay and Weber 2008).

To estimate the geographical distribution of the price density, multiple methods such as multiple regression (McCluskey et al. 2000) or Bayesian smoothing (Clapp, Kim, and Gelfand 2002) have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. This problem is known as ``finite area smoothing'' and occurs when predictions from a model are approximated across geographical barriers, such as irregular coastal shapes, which would lead to poorly predicted values (Ramsay 2002). To deal with the influence of geographical boundaries on price estimation, a generalised additive model using soap film smoothing has been used (Wood, Bravington, and Hedley 2008).

@@

OpenStreetMap combines information about more than 400 urban features including road information and building information to categorize features such as amenities, leisure or tourism structures (OpenStreetMap 2020a). Among these features, only 160 are available or relevant to the Dublin Area. Urban features are obtained using the R package osmdata (Padgham et al. 2017) which queries the Overpass API (OpenStreetMap 2020b). The distance between each property and the closest point corresponds to each of the 160 urban features.

@@

In addition to the distance to every urban feature, each property sold in 2018 was related to 48 socio-economic features of the small area including the property, as measured by the Irish census 2016. The results of Irish 2016 census consultation is accessible through both the Central Statistics Office (Central Statistics Office 2020) and All-Island Research Observatory (All-Island Research Observatory 2020). The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

\hypertarget{extreme-gradient-boosting}{%
\subsection{eXtreme Gradient Boosting}\label{extreme-gradient-boosting}}

Lastly, an eXtreme Gradient Boosting (XGBoost) model was calculated to examine the extent to which property prices rely on urban and on socio-economic features. The XGBoost algorithm optimizes the prediction accuracy by performing boosting iterations (set to 100 iterations as the best trade-off between accuracy and speed), thereby minimizing the root mean squared error when the model's objective is a continuous output (Friedman 2001). The R package ``xgboost'' (v.1.1.1.1) is used to implement the XGBoost algorithm (Chen and Guestrin 2016). With this version, two types of booster are available: tree based and linear regression based. Different types of learning task can also be defined according to the distribution of the output. In the case of a continuous output, they include regression with squared loss, regression with squared log loss, gamma regression with log-link, and tweedie regression with log-link.

The original dataset has been randomly split to 80\% for training and 20\% for testing the models. The prediction accuracy is estimated using the coefficient of determination (\(R^2\)), the Root Mean Square Error (\(RMSE\)) and the Mean Absolute Error (\(MAE\)) of the predicted property price values of the test dataset.

Both tree and linear regression based boosting model are used for every of the continuous output learning task in order to compare their accuracy with urban and socio-economic features. As no specific weight or bias in the models has been hypothesized, the parameters of the XGBoost function have been set to the default value. Then, the most accurate model is used to identify the features that are contributing the most to the prediction of property prices. This importance is a measure of the improvement in accuracy brought by a feature based on its coefficient's absolute magnitude.

@@

In 2018, the average price of a sold property in Dublin was €384,892 with a standard deviation of €175,476. In order to remove potential human errors and outliers, prices higher or lower than 1 SD were removed from the original dataset.

\begin{figure}[!h]
\includegraphics[width=0.98\columnwidth]{manuscript_files/figure-latex/distrib-plot-1} \caption{Overall distribution of the property prices density in the city of Dublin in 2018}\label{fig:distrib-plot}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=0.98\columnwidth]{manuscript_files/figure-latex/distrib-gps-1} \caption{Geographic distribution of the property prices density in the city of Dublin in 2018 using a generalized additive model with soap film smooth parameter to take into account the influence of coastal boundaries. Estimates outside of the 95\% Confidence Interval were removed.}\label{fig:distrib-gps}
\end{figure}

The density of housing prices reveals a unimodal distribution of property prices (Figure \ref{fig:distrib-plot}). In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \ref{fig:distrib-gps}).

@@

In order to compare the possibilities of the different type of parameters in the XGBoost models,
all combinations of tree and linear regression based boosters with all possible output objectives (i.e squared error, squared log error, gamma, and tweedie) are displayed Table \ref{tab:osm-model-comparison}.

\begin{table}[H]

\caption{\label{tab:osm-model-comparison}Accuracy of XGBoost models using urban features according the type of gradient boosting method and output objective.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrl}
\toprule
Objective & $RMSE$ & $R^2$ & $MAE$\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Tree based}}\\
\hspace{1em}squarederror & €127,783 & 0.44 & €85,115\\
\hspace{1em}squaredlogerror & €426,989 & - & €391,670\\
\hspace{1em}gamma & €129,052 & 0.43 & €87,039\\
\hspace{1em}tweedie & €126,626 & 0.45 & €85,073\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Linear regression based}}\\
\hspace{1em}squarederror & €150,833 & 0.21 & €111,788\\
\hspace{1em}squaredlogerror & €406,947 & 0.00 & €369,713\\
\hspace{1em}gamma & €180,313 & 0.06 & €136,059\\
\hspace{1em}tweedie & €172,903 & 0.08 & €130,148\\
\bottomrule
\multicolumn{4}{l}{\textit{Note: }}\\
\multicolumn{4}{l}{Missing value ares due to incompatibilities between data and methods.}\\
\end{tabular}
\end{table}

The model comparison reveals that tree based boosting models are more accurate than linear regression based model. In addition, within the possible objectives of the tree based boosting models, the tweedie regression with log-link appears to obtain the highest \(R^2\) value as well as the lowest \(RMSE\) and \(MAE\) prediction errors.

Using the XGBoost algorithm with a tree based booster and a tweedie regression with log-link learning objective, the 160 urban features are explaining 45\% of the property price variance (\(F(1, 2076) = 1,690.80\), \(p < .001\)) with a \(RMSE\) of €126,626 and a \(MAE\) of €85,073.07.

The overall correlation of the predicted property prices with urban features is shown in Figure \ref{fig:osm-features-xgb}A. It can be noticed that property prices situated at the low and the high end of the range are the most difficult to predict (Figure \ref{fig:osm-features-xgb}B). A possible explanation is the absence of distinctive and recurrent patterns in urban features for these properties. However, the prices higher than €300,000 are potentially driving down the prediction accuracy due to outliers.

\begin{figure}[!h]
\includegraphics[width=0.98\columnwidth]{manuscript_files/figure-latex/osm-features-xgb-1} \caption{Property price prediction accuracy (A) and prediction error (B) using urban features with XGBoost.}\label{fig:osm-features-xgb}
\end{figure}

By analysing their importance (Table \ref{tab:osm-features-table}), the most relevant geographical features to predict housing prices are the distance to an embassy (15.6\%) and the distance to natural grasslands such as parks and gardens (4.8\%).

\begin{table}[H]

\caption{\label{tab:osm-features-table}Top 10 most important urban features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
amenity & embassy & 15.6\%\\
natural & grassland & 4.8\%\\
power & line & 2.4\%\\
route & bus & 2.4\%\\
amenity & bar & 2.0\%\\
boundary & political & 1.9\%\\
place & island & 1.6\%\\
boundary & administrative & 1.5\%\\
highway & residential & 1.4\%\\
barrier & wall & 1.4\%\\
\bottomrule
\end{tabu}
\end{table}

@@

Similarly to urban features, the accuracy of different XGBoost models using socio-economic features to predict property prices are displayed Table \ref{tab:census-model-comparison}.

\begin{table}[H]

\caption{\label{tab:census-model-comparison}Accuracy of XGBoost models using socio-economic features according the type of gradient boosting method and output objective.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrl}
\toprule
Objective & $RMSE$ & $R^2$ & $MAE$\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Tree based}}\\
\hspace{1em}squarederror & €131,270 & 0.42 & €86,892\\
\hspace{1em}squaredlogerror & €426,989 & - & €391,670\\
\hspace{1em}gamma & €129,221 & 0.43 & €86,949\\
\hspace{1em}tweedie & €130,927 & 0.42 & €86,707\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Linear regression based}}\\
\hspace{1em}squarederror & €146,781 & 0.26 & €107,890\\
\hspace{1em}squaredlogerror & €406,108 & 0.01 & €368,877\\
\hspace{1em}gamma & €182,432 & 0.08 & €136,441\\
\hspace{1em}tweedie & €164,439 & 0.13 & €122,259\\
\bottomrule
\multicolumn{4}{l}{\textit{Note: }}\\
\multicolumn{4}{l}{Missing value ares due to incompatibilities between data and methods.}\\
\end{tabular}
\end{table}

The comparison of accuracy metrics reveals that tree based boosting models obtain better results than linear regression based boosting models. Within these tree based boosting models, the gamma regression with log-link is the best model as it obtains the highest \(R^2\) and lowest \(RMSE\) values.

\begin{figure}[!h]
\includegraphics[width=0.98\columnwidth]{manuscript_files/figure-latex/census-features-xgb-1} \caption{Property price prediction accuracy (A) and prediction error (B) using socio-economic features with XGBoost.}\label{fig:census-features-xgb}
\end{figure}

Using the XGBoost algorithm with a tree based booster and a gamma regression with log-link learning objective, the 48 socio-economic features explain 43\% of the property price variance (\(F(1, 2076) = 1,554.37\), \(p < .001\)) with a \(RMSE\) of €129,221 and a \(MAE\) of €86,949.15.

@@

According to the analysis of the relative importance of socio-economic features (Table \ref{tab:census-features-table}), the most important socio-economic features is the proportion individual not providing regular unpaid personal help for a friend or family member with a long-term illness, health problem or disability (32.8\%). It also appears that the proportion of large properties (i.e., houses with eight or more rooms) in the small area containing the property (19.1\%) as well as the proportion of people reporting that they have no religion (4.2\%), influence the model.

\begin{table}[H]

\caption{\label{tab:census-features-table}Top 10 most important socio-economic features contributing to the property price prediction.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Feature Category & Feature Type & Importance\\
\midrule
carers & Provides No Care & 32.8\%\\
housing rooms & 8 or more Rooms & 19.1\%\\
religion & No Religion & 4.2\%\\
population & Age 0 - 14 & 3.2\%\\
housing rooms & 6 Rooms & 2.1\%\\
carers & 1-19 hours unpaid PW & 1.9\%\\
housing tenure & Social Rented & 1.8\%\\
general health & Very Good & 1.7\%\\
religion & Other Catholic & 1.6\%\\
general health & Good & 1.5\%\\
\bottomrule
\end{tabu}
\end{table}

@@

While Dublin is a very specific city due to its low population density spread on a large surface, its housing market is one of the most expensive in Europe. The main reason is the structure of the market, mainly made of residential properties and few apartment buildings. In this study, the prices of 10387 properties sold in 2018 in the area of Dublin were analysed from the Property Services Regulatory Authority dataset in order to identify the potential factors correlating with these prices. The enforcement of a housing price registry has been shown as being an essential tool to regulate the housing market (Kitchin 2013). However, a major limitation of this study is the absence of characteristics for the property itself. Indeed, prices are mainly determined by factors such as the size of the property or the number of rooms. In addition, while some features appear to be more important than others in their contribution to property prices, these results are not causal. Thus, features' importance can be either the cause or the consequence of property prices in a neighborhood.

Despite the absence of property characteristics, urban and socio-economic features were used to identify spatial correlates to the housing prices. Results revealed that proximity to embassies and grasslands is a driver of property prices. Embassies are in general located in the most expensive areas of cities which are also the most aesthetically pleasing and the most secure part of cities. In addition, the density of embassies and properties allocated to embassy staff reduces the density of property available on the market and drives their prices up. The distance to a park or a natural area is also a very important factor for property prices. Results also revealed that the density of individual not providing regular unpaid personal help in the area is the most important socio-economic feature to predict property prices. This feature is related to the multiple factors such as the average age of the household or the economic status of the household as they are related to general health (Smith 1999). Large properties in the area and the proportion of individuals reporting having no religion are also relatively important. Again, the size of the property sold was not included in the dataset; however, the higher the density of large properties in the area, the higher the probability for the property to be a large property. Finally, the relation with the expression of religious belief, or more precisely its absence, is a very interesting feature. Again, the relationship between religion and income is not confirmed by the literature. Whereas some studies reveal a positive correlation between religion and income (Guiso, Sapienza, and Zingales 2003; Elgin et al. 2013) others reveals no evidence of such relationship (De La O and Rodden 2008). Here, it appears that prices of properties are higher in areas with a high density of people indicating that they have no religion.

@@

The evolution of housing prices is a pressing issue in most European capitals and specially in Dublin. Given their significant increase, properties are less and less affordable for individuals. By preforming a feature analysis with urban and socio-economic features, it is possible to evaluate and predict the potential price of a property. Indeed features such as the presence of embassies or parks are criteria that influence significantly the price of properties. Similarly, the characteristics of inhabitants in the area such as religion, health and age is correlated to the evolution of housing prices. These results allow an understanding of why some areas have higher prices than others which is a relevant information not only for real estate agents in charge of property valuations but also to buyers in order to approximate the real value of a property.

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-airo2020}{}%
All-Island Research Observatory. 2020. ``AIRO Data Store.'' June 1, 2020. \url{http://airo.maynoothuniversity.ie/datastore}.

\leavevmode\hypertarget{ref-antonakakis2016dynamic}{}%
Antonakakis, Nikolaos, and Christos Floros. 2016. ``Dynamic Interdependencies Among the Housing Market, Stock Market, Policy Uncertainty and the Macroeconomy in the United Kingdom.'' \emph{International Review of Financial Analysis} 44: 111--22.

\leavevmode\hypertarget{ref-aragones2007neural}{}%
Aragonés, José R, Carlos Blanco, and Pablo García Estévez. 2007. ``Neural Network Volatility Forecasts.'' \emph{Intelligent Systems in Accounting, Finance \& Management: International Journal} 15 (3-4): 107--21.

@@

@@

\leavevmode\hypertarget{ref-cso2020}{}%
Central Statistics Office. 2020. ``Census 2016 Small Area Population Statistics.'' June 1, 2020. \url{https://www.cso.ie/en/census/census2016reports/census2016smallareapopulationstatistics/}.

@@

\leavevmode\hypertarget{ref-nominatim2020}{}%
Hoffmann, Sarah. 2020. ``Nominatim: Open-Source Geocoding with Openstreetmap Data.'' June 1, 2020. \url{https://nominatim.org/}.

\leavevmode\hypertarget{ref-kitchin2013making}{}%
Kitchin, Rob. 2013. ``Making Informed Decisions on Future Housing Policy.'' \emph{Housing Ireland: A Journal for Irish Housing Professionals}, no. 3: 36--41.

@@

\leavevmode\hypertarget{ref-rppr2020}{}%
Property Services Regulatory Authority. 2020. ``Residential Property Price Register.'' June 1, 2020. \url{https://propertypriceregister.ie}.
